---
title: "Activity Analysis"
author: "Stephen Hobbs"
date: "April 5, 2016"
output: html_document
---
Download the data set from this url: https://www.coursera.org/learn/reproducible-research/peer/gYyPt/course-project-1
Click on the hyperlink, "Activity monitoring data."
Download and save to your hard drive. Assign the variable, data, to the file.
The variables included in this dataset are:

    steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
    date: The date on which the measurement was taken in YYYY-MM-DD format
    interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.
Loading and preprocessing the data, import .csv file.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/stephenhobbs1/Data science/5-Reproducible Research")
data <- read.csv("activity.csv", stringsAsFactors=FALSE, header = TRUE)
```
Load R packages to be used later.
```{r, echo=TRUE}
library(plyr)
library(lattice)
```
What is mean total number of steps taken per day?
Summarize the data by date and number of steps using function ddpy from plyr package. Remove "NA" values.
Assign this data set to "tmp" variable. "tmp" gives you the total number of steps by date.
```{r echo=TRUE}
#Summarize dataframe by date, total steps
tmp<-ddply(data,.(date),summarise, totalSteps=sum(steps, na.rm=TRUE)) 
# Sum of total steps, 570,608 steps.
sum(tmp$totalSteps) 
```
Here is the histogram.
```{r histogram, echo=TRUE}
hist(tmp$totalSteps, breaks = 10, xlab = "Total Steps",
     main = "Histogram of Total Steps Taken Each Day")
```


Next is the code of the average and median steps. 
```{r echo=TRUE}
# Average step is 9354.23
mean(tmp$totalSteps, na.rm=TRUE) 
# Median step is 10,395
median(tmp$totalSteps, na.rm=TRUE) 
```
What is the average daily activity pattern?
Use the plyr package, ddply function, to get the average steps per time interval and plot. Assign new variable, tmp2, to this data frame.

```{r echo=TRUE}
#Summarize date set by date, average steps/interval
tmp2<-ddply(data,.(interval),summarise, AverageSteps=mean(steps, na.rm=TRUE)) 
# Average number of steps to use in imputing NA values
mean(tmp2$AverageSteps) 
# Plot average steps per interval using a line as the type. 
plot(tmp2, type="l", main = "Average steps per interval", ylab="Average Steps", xlab="Interval, minute units")
```
 
 Which interval had the most steps? Run this code to discover.
```{r echo=TRUE}
#Answer is 206.17 steps at 0835.
tmp2[which.max(tmp2$AverageSteps),]
```
Imputing missing values
I took the average of the steps per interval and substituted the average for all NA values.
Number of NA values is 2304. Here is the code.
``` {r echo= TRUE}
# Code shows number of "NA" values, or 2304 "NA" values.
sum(is.na(data)) 
```
Then we want to group data on date and total steps. First, we calculate the average of the 5 minute interval averages. This will then be the number we use to impute the value for NA. That number is 37.3826.
```{r echo=TRUE}
#Replacing missing NA value with average of average of 5 min values. Merge original data set with tmp2, which shows average.
datanew<-merge(data, tmp2, by="interval", sort=TRUE) 
index <- is.na(datanew$steps) #Returns logical vector, true = missing value
#Here we impute. Take out rows of missing values and replace with #AverageSteps where there was NA, missing value
datanew$steps[index] <- datanew$AverageSteps[index] 
#Group on date and total steps
tmp3<-ddply(datanew, .(date), summarise, totalsteps=sum(steps, na.rm = TRUE)) 
```

Below is the code for a histogram of averages with using imputed average for missing values.
``` {r echo=TRUE }
hist(tmp3$totalsteps, breaks=10, main = "Histogram of total steps taken each day, Impute NA", xlab="Total Steps") #Histogram with imputed missing values
```

Below is the code for mean and median steps with imputed average replacing missing value. The mean and median wer 10,766.19 steps. The average and median using imputed average in place of NA differ from the average and median with NA in the data set, 9534 and 10,395 respectively.
``` {r echo = TRUE}
# Imputed mean of steps, NA removed. Average steps = 10,766.19 steps
mean(tmp3$totalsteps) 
# Imputed median of steps, NA removed. Median steps  = 10,766.19 steps
median(tmp3$totalsteps)
```
Differences between weekday and weekend activity
 we separate steps of weekdays from weekends. 
``` {r echo = TRUE}
# Convert dates into weekdays and weekends
tmp4<-transform(data, date = strptime(date, format="%Y-%m-%d"))
#Use weekdays() function to classify day.
weekdays(tmp4$date) %in% c("Saturday", "Sunday") 
# Set logical if day is weekday or weekend.
factor(weekdays(tmp4$date) %in% c("Saturday", "Sunday"), level=c(TRUE, FALSE), labels=c("weekend", "weekday"))
# Assign factor variable to day
day1<-factor(weekdays(tmp4$date) %in% c("Saturday", "Sunday"), level=c(TRUE, FALSE), labels=c("weekend", "weekday"))
#Create new column, weekend, weekend using mutate from plyr
tmp4<-mutate(tmp4, day=day1)
#Plot weekend and weekday averages; change day from POSIXlt format to character format because ddply does not work on posixlt; it only works on charactoer & number vector.
tmp4<-transform(tmp4, day1=as.character((day1)))  
tmp5<-ddply(tmp4,.(interval,day1), summarize, averagesteps=mean(steps, na.rm = TRUE))
#Plot weekday and weekend average steps per interval using line graph.
xyplot(averagesteps~interval|day1, data = tmp5, 
       ylab="Average Steps", 
       type = "l",
       main="Comparison of average steps between weekday and weekend")
```

Weekend average is higher on average but max value was highest during weekday.
